MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-08-01 MAESTROeX_I/algorithm.tex
minor changes and comments through the alg description

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry.
The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the base state cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the base state cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels. Reprinted from Paper V \citep{MAESTRO_V}. }
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of constant radius of a Cartesian grid variable onto a one-dimensional base state.
In planar geometries, this is a straightforward arithmetic average of values in cells at
a particular height since the base state cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any 
there-dimensional Cartesian cell-center can map to.  
Specifically, for every three-dimensional Cartesian cell, there exists an integer $m$ such that the radius to the center of the star is given by 
\begin{equation}
\hat{r}_m=\Delta x\sqrt{0.75+2m}.\label{eqn:radii}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.5in]{./figs/base_spherical_new}
\caption{\label{Fig:NewBaseGrid}  
A direct mapping between the base state cell centers (red squares) and the Cartesian grid cell centers (blue crosses) 
is enforced by computing the average of the grid cell centers that share the same radial distance from 
the center of the star.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Figure \ref{Fig:NewBaseGrid} is a two-dimensional illustration 
(two dimensions is chosen in the figure for ease of exposition; this mapping is only used for three-dimensional spherical stars) 
of the relationship between the Cartesian grid state and the one-dimensional base state array.
We compute the lateral average by first summing the values in all the cells associated with each radius,
dividing by the number of contributing cells to obtain the arithmetic average, and then use quadratic interpolation to map this data onto a one-dimensional base state.
Previously, for spherical problems MAESTRO only allowed for a base state with constant $\Delta r$ (typically equal to $\Delta x/5$).

@@

In this paper we explore a new option to retain an irregularly-spaced base state to eliminate mapping errors from the ``fill'' module.
For the lateral average, as before we first sum the values in all the cells associated with each radius and divide
by the number of contributing cells to obtain the arithmetic average.  However we do not map this onto a uniformly spaced
base state and retain the use of this irregularly-spaced base state.
The advantage with this approach is that the fill module does not require any interpolation.
A potential benefit to eliminating the mapping error is to consider a spherical star in hydrostatic equilibrium at rest.
In the absence of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors 
in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error 
from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced 
array, the effects due to interpolation error are nearly eliminated, but not completely eliminated since there are still 
machine precision effects resulting from averaging a large number of numerical values.
We note that $\Delta r$ decreases as the base state moves further from the center of the star, 
which results in far more total cells in the irregularly-spaced array than the previous uniformly-spaced array.

\subsection{Temporal Integration Scheme}\label{Sec:Temporal Integration Scheme}
We now describe the new temporal integration scheme, which we note can be used for the original base state mapping (with uniform base state grid spacing) as well as the new irregularly spaced base state mapping.
Previously we adopted an approach where we split the velocity into a base state component, $w_0(r,t)$, 
and a local velocity $\Ubt(\xb,t)$, so that
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r, \label{eq:velsplit}
\end{equation}
where $\eb_r$ is the normal vector in the outward radial direction.
We used $w_0$ to provide an estimate of the base state density evolution over a time step.
This resulted in some unnecessary complications to the temporal integration scheme.
Our new temporal integration scheme uses full velocities to evolve advection and reactions instead, and only uses the above splitting to satisfy the velocity divergence constraints due to boundary considerations. 
This produces a cleaner scheme than the one from Paper V since we use the velocity directly rather than more complex terms involving the perturbational velocity.
Additionally, the new scheme uses a simpler predictor-corrector approach to the base state density and pressure that no 
longer requires evolution equations and the associated numerical discretizations to update the base state, greatly 
simplifying the algorithm while retaining the same overall second-order accuracy in time.

At the beginning of each time step we have the cell-centered Cartesian grid state,
$(\Ub,\rho X_k,\rho h)^n$, and nodal Cartesian grid state, $\pi^{n-\myhalf}$, and base state $(\rho_0,p_0)^n$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution; see \cite{XRB_I}.  We are still exploring this option for spherical stars.}, e.g.,
$T = T(\rho,p_0,X_k)$, where $p_0$ has been mapped to the Cartesian grid using the fill module,
and ($\gammaonebar,\beta_0)$ are similarly computed from $(\rho,p_0,X_k)$;
see Appendix A of Paper I and Appendix C of Paper III for details on how these terms are computed.

The overall flow of the algorithm begins with a second-order Strang splitting approach to integrate the advection-reaction system for 
the thermodynamic variables $(\rho X_k, \rho h)$, followed by a second-order projection methodology to integrate the velocities subject to a divergence constraint.  Within the thermodynamic variable update we use a predictor-corrector approach to achieve second-order accuracy in time.
To summarize:
\begin{itemize}
\item In {\bf Step 1} we react the thermodynamic variables over the first $\Delta t/2$ interval.
\item In {\bf Steps 2--4} we advect the thermodynamic variables over $\Delta t$.  Specifically, we compute an estimate for the expansion term, $S$, compute face-centered, time-centered velocities that satisfy the divergence constraint, and then advect the thermodynamic variables.
\item In {\bf Step 5} we react the thermodynamic variables over the second $\Delta t/2$ interval\footnote{After this step we could skip to the velocity advance in {\bf Steps 10--11}, however the overall scheme would be only first-order in time, so {\bf Steps 6-9} can be thought of as a trapezoidal corrector step.}.
\item In {\bf Steps 6--8} we redo the advection in {\bf Steps 2--4} but are able to use the trapezoidal rule to time-center certain quantities such as $S$, $\rho_0$, etc.
\item In {\bf Step 9} we redo the reactions from {\bf Step 5} beginning with the improved results from {\bf Steps 6--8}.
\item In {\bf Steps 10--11} we advect the velocity, and then project these velocities so they satisfy the divergence constraint.
\end{itemize}

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell-edge $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ integrates the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outputs are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the time step that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

In stratified low Mach number models, due to the extreme variation in density, the velocity can become very large in low density regions at the edge of the star.
Throughout Papers II-V, we have employed two techniques to help control these dynamics without significantly affecting the dynamics in the convective region.
First, we use a cutoff density technique, where we hold the density constant outside a specified radius (typically near where the density is $\sim$4 orders of magnitude smaller than the largest densities in the simulation).
Second, we employ a sponge technique where we artificially damp the velocities near and beyond the cutoff region.
For more details, refer to Paper V and the previous references cited within.

>>>>

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry.
The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the base state cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the base state cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels. Reprinted from Paper V \citep{MAESTRO_V}. }
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of constant radius of a Cartesian grid variable and stores the result in a one-dimensional base state array.
In planar geometries, this is a straightforward arithmetic average of values in cells at
a particular height since the base state cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any 
there-dimensional Cartesian cell-center can map to.  
Specifically, for every three-dimensional Cartesian cell, there exists an integer $m$ such that the distance
from the cell center to the center of the star is given by 
\begin{equation}
\hat{r}_m=\Delta x\sqrt{0.75+2m}.\label{eqn:radii}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.5in]{./figs/base_spherical_new}
\caption{\label{Fig:NewBaseGrid}  
A direct mapping between the base state cell centers (red squares) and the Cartesian grid cell centers (blue crosses) 
is enforced by computing the average of the grid cell centers that share the same radial distance from 
the center of the star.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Figure \ref{Fig:NewBaseGrid} is a two-dimensional illustration 
(two dimensions is chosen in the figure for ease of exposition; this mapping is only used for three-dimensional spherical stars) 
of the relationship between the Cartesian grid state and the one-dimensional base state array.
We compute the lateral average by first summing the values in all the cells associated with each radius,
dividing by the number of contributing cells to obtain the arithmetic average, and then use quadratic interpolation to map this data onto a one-dimensional base state.
Previously, for spherical problems MAESTRO only allowed for a base state with constant $\Delta r$ (typically equal to $\Delta x/5$).

@@

In this paper we explore a new option to retain an irregularly-spaced base state to eliminate mapping errors from the ``fill'' module.
For the lateral average, as before we first sum the values in all the cells associated with each radius and divide
by the number of contributing cells to obtain the arithmetic average.  However we do not interpolate this onto a uniformly spaced
base state and retain the use of this irregularly-spaced base state.
The advantage with this approach is that the fill module does not require any interpolation.
A potential benefit to eliminating the mapping error is to consider a spherical star in hydrostatic equilibrium at rest.
In the absence of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors 
in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error 
from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced 
array, the effects due to interpolation error are nearly eliminated, but not completely eliminated since there are still 
machine precision effects resulting from averaging a large number of numerical values.
We note that $\Delta r$ decreases as the base state moves further from the center of the star, 
which results in far more total cells in the irregularly-spaced array than the previous uniformly-spaced array.

\subsection{Temporal Integration Scheme}\label{Sec:Temporal Integration Scheme}
We now describe the new temporal integration scheme, noting that it can be used for the original base state mapping 
(with uniform base state grid spacing) as well as the new irregularly spaced base state mapping.
Previously we adopted an approach where we split the velocity into a base state component, $w_0(r,t)$, 
and a local velocity $\Ubt(\xb,t)$, so that
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r, \label{eq:velsplit}
\end{equation}
where $\eb_r$ is the normal vector in the outward radial direction.
We used $w_0$ to provide an estimate of the base state density evolution over a time step.
This resulted in some unnecessary complications to the temporal integration scheme including
base state advection modules for density, enthalpy, and velocity, as well as more 
cumbersome split velocity dynamics evolution equations.
Our new temporal integration scheme uses full velocities for scalar and velocity advection,
and only uses the above splitting to satisfy the velocity divergence constraint due to boundary considerations
at the edge of the star.
This results in a much simpler numerical scheme than the one from Paper V 
since we use the velocity directly rather than more complex terms involving the perturbational velocity.
Additionally, the new scheme uses a simpler predictor-corrector approach to the base state density and pressure that no 
longer requires evolution equations and numerical discretizations to update the base state, greatly 
simplifying the algorithm while retaining the same overall second-order accuracy in time.

At the beginning of each time step we have the cell-centered Cartesian grid state,
$(\Ub,\rho X_k,\rho h)^n$, and nodal Cartesian grid state, $\pi^{n-\myhalf}$, and base state $(\rho_0,p_0)^n$.
At any time, the associated density, composition, and enthalpy can be trivially computed using, e.g.,
\begin{equation}
\rho^n = \sum_k(\rho X_k)^n, \quad
X_k^n = (\rho X_k)^n / \rho^n, \quad
h^n = (\rho h)^n / \rho^n.
\end{equation}
Temperature is computed using the equation of state\footnote{As described in Paper V, for planar problems we compute temperature using $h$ instead of $p_0$, since we have successfully developed volume discrepancy schemes to effectively couple the enthalpy to the rest of the solution; see \cite{XRB_I}.  We are still exploring this option for spherical stars.}, e.g.,
$T = T(\rho,p_0,X_k)$, where $p_0$ has been mapped to the Cartesian grid using the fill module,
and ($\gammaonebar,\beta_0)$ are similarly computed from $(\rho,p_0,X_k)$;
see Appendix A of Paper I and Appendix C of Paper III for details on how $\beta_0$ is computed.

The overall flow of the algorithm begins with a second-order Strang splitting approach to integrate the advection-reaction system for 
the thermodynamic variables $(\rho X_k, \rho h)$, followed by a second-order projection methodology to integrate the velocities subject to a divergence constraint.  Within the thermodynamic variable update we use a predictor-corrector approach to achieve second-order accuracy in time.
To summarize:
\begin{itemize}
\item In {\bf Step 1} we react the thermodynamic variables over the first $\Delta t/2$ interval.
\item In {\bf Steps 2--4} we advect the thermodynamic variables over $\Delta t$.  Specifically, we compute an estimate for the expansion term, $S$, compute face-centered, time-centered velocities that satisfy the divergence constraint, and then advect the thermodynamic variables.
\item In {\bf Step 5} we react the thermodynamic variables over the second $\Delta t/2$ interval\footnote{After this step we could skip to the velocity advance in {\bf Steps 10--11}, however the overall scheme would be only first-order in time, so {\bf Steps 6-9} can be thought of as a trapezoidal corrector step.}.
\item In {\bf Steps 6--8} we redo the advection in {\bf Steps 2--4} but are able to use the trapezoidal rule to time-center certain quantities such as $S$, $\rho_0$, etc.
\item In {\bf Step 9} we redo the reactions from {\bf Step 5} beginning with the improved results from {\bf Steps 6--8}.
\item In {\bf Steps 10--11} we advect the velocity, and then project these velocities so they satisfy the divergence constraint while updating $\pi$.
\end{itemize}

There are a few key numerical modules we use in each time step.
\begin{itemize}
\item {\bf Average}$[\phi]\rightarrow[\overline\phi]$ computes the lateral average of a quantity over a layer at constant radius $r$, as described above in Section \ref{Sec:Spatial}.
\item {\bf Enforce HSE}$[\rho_0]\rightarrow[p_0]$ computes the base state pressure, $p_0$, from a base state density, $\rho_0$ by integrating the hydrostatic equilibrium condition in one dimension. 
This follows equation (A10) in Paper V, noting that for the irregularly spaced base state case, $\Delta r$ is not constant, where $\Delta r_{j+1/2} = r_{j+1} - r_j$ for cell-edge $j+1/2$.  
The base state pressure remains equal to a constant value at the location of a prescribed cutoff density outward for the entire simulation.
\item {\bf React State}$[(\rho X_k)^{\rm in},(\rho h)^{\rm in},p_0]\rightarrow[(\rho X_k)^{\rm out},(\rho h)^{\rm out},(\rho\dot\omega),(\rho\Hnuc)]$ 
uses VODE \citep{vode} to integrate the species and enthalpy due to reactions over $\Delta t/2$ by solving
\begin{equation}
\frac{dX_k}{dt} = \dot\omega_k(\rho,X_k,T); \qquad
\frac{dT}{dt} = \frac{1}{c_p}\left(-\sum_k\xi_k\dot\omega_k + H_{\rm nuc})\right)
\end{equation}
The inputs are the species, enthalpy, and base state pressure, and the outputs are the species, enthalpy, reaction rates, and nuclear energy generation rate.
See Paper III for details.
\end{itemize}
Each time step is constrained by the standard advective CFL condition,
\begin{equation}
\Delta t = \sigma^{\rm CFL} \min_i(\Delta x / U_i),
\end{equation}
where for our simulations we typically use $\sigma^{\rm CFL}\sim 0.7$ and the minimum is taken over all spatial directions over all cells.
There are additional constraints on the time step that are typically much less restrictive than the advective CFL including
the acceleration due to the buoyancy force (sometimes in effect when the velocity is approximately zero at the start of some simulations) 
and the local magnitude of the divergence constraint (to prevent too much mass evacuation from a cell in a time step); see Section 3.4 in Paper III for details.

In stratified low Mach number models, due to the extreme variation in density, the velocity can become very large in low density regions at the edge of the star.
These large velocities can severely affect the time step, so
throughout Papers II-V, we have employed two techniques to help control these dynamics without significantly 
affecting the dynamics in the convective region.
First, we use a cutoff density technique, where we hold the density constant outside a specified radius (typically near where the density is $\sim$4 orders of magnitude smaller than the largest densities in the simulation).
Second, we employ a sponge technique where we artificially damp the velocities near and beyond the cutoff region.
For more details, refer to Paper V and the previous references cited within.

### 2019-08-01 MAESTROeX_I/equations.tex
minor changes and comments through the alg description

\section{Governing Equations}\label{sec:equations}
Low Mach number models for reacting flow were originally derived using asymptotic analysis
\citep{rehm1978equations,majda1985derivation} and used in terrestrial combustion applications
\citep{knio1999semi,day2000numerical}.  These models have been extended to nuclear flames
in astrophysical environments using adaptive algorithms in space and time \citep{Bell:2004}.
In Papers I-III, we extended this work and the atmospheric model by \citet{durran:1989} by deriving a model and algorithm suitable for stratified astrophysical flow.
We take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field, as explained below.
The resulting model is a series of evolution equations for mass, momentum, and energy, subject
to to an additional constraint on velocity.  The evolution equations are
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc.\label{eq:enthalpy}
\end{eqnarray}
Here $\rho$, $\Ub$, and $h$ are the mass density,  
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$ and energy release per time per unit mass $\Hnuc$.
The species are constrained such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
The total pressure is decomposed into a one-dimensional hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that 
$p = p_0 + \pi$ and $|\pi|/p_0 = \mathcal{O}({\rm Ma}^2)$ (we use $\xb$ to represent the Cartesian coordinate 
directions of the full state and $r$ to represent the radial coordinate direction for 
the base state).  
By comparing the momentum equation (\ref{eq:momentum}) to the momentum equation used in equation (2) in Paper V, we
note that we are using a formulation that enforces conservation of total energy in the 
low Mach number system in the absence of external heating or viscous terms \citep{kleinpauluis,Vasil2013}.
We have previously validated this approach in modeling sub-Chandrasekhar white dwarfs using MAESTRO \citep{subChandra_II}. 
We also define a one-dimensional base state density, $\rho_0 = \rho_0(r,t)$, that represents the lateral average (see Section \ref{Sec:Spatial}) of $\rho$ and is in hydrostatic equilibrium with $p_0$, i.e., 
\begin{equation}
\nabla p_0 = -\rho_0 g\eb_r, \label{eq:HSE}
\end{equation}
where $g=g(r,t)$ is the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the outward radial direction. 
Here $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho) |_s$ (evaluated with entropy, $s$, constant).
Thermal diffusion is not discussed in this paper, but we have previously described the modifications to the original algorithm required 
for implicit thermal diffusion in \cite{XRB_I}; inclusion of these effects in the new algorithm presented here is straightforward.

>>>>

\section{Governing Equations}\label{sec:equations}
Low Mach number models for reacting flow were originally derived using asymptotic analysis
\citep{rehm1978equations,majda1985derivation} and used in terrestrial combustion applications
\citep{knio1999semi,day2000numerical}.  These models have been extended to nuclear flames
in astrophysical environments using adaptive algorithms in space and time \citep{Bell:2004}.
In Papers I-III, we extended this work and the atmospheric model by \citet{durran:1989} by deriving a model and algorithm suitable for stratified astrophysical flow.
We take the standard equations of reacting, compressible flow, and recast the equation
of state (EOS) as a divergence constraint on the velocity field.
The resulting model is a series of evolution equations for mass, momentum, and energy, subject
to an additional constraint on velocity.  The evolution equations are
\begin{eqnarray}
\frac{\partial(\rho X_k)}{\partial t} &=& -\nabla\cdot(\rho X_k\Ub) + \rho\omegadot_k,\label{eq:species}\\
\frac{\partial\Ub}{\partial t} &=& -\Ub\cdot\nabla\Ub  - \frac{\beta_0}{\rho}\nabla\left(\frac{\pi}{\beta_0}\right) - \frac{\rho-\rho_0}{\rho} g\eb_r,\label{eq:momentum}\\
\frac{\partial(\rho h)}{\partial t} &=& -\nabla\cdot(\rho h\Ub) + \frac{Dp_0}{Dt} + \rho\Hnuc.\label{eq:enthalpy}
\end{eqnarray}
Here $\rho$, $\Ub$, and $h$ are the mass density,
velocity and specific enthalpy, respectively, and
$X_k$ are the mass fractions of species $k$ with associated
production rate $\omegadot_k$ and energy release per time per unit mass $\Hnuc$.
The species are constrained such that $\sum_k X_k = 1$ giving $\rho = \sum_k (\rho X_k)$ and
\begin{equation}
\frac{\partial\rho}{\partial t} = -\nabla\cdot(\rho\Ub).
\end{equation}
The total pressure is decomposed into a one-dimensional hydrostatic base state
 pressure, $p_0 = p_0(r,t)$, and a dynamic pressure, $\pi = \pi(\xb,t)$, such that
$p = p_0 + \pi$ and $|\pi|/p_0 = \mathcal{O}({\rm Ma}^2)$ (we use $\xb$ to represent the Cartesian coordinate
directions of the full state and $r$ to represent the radial coordinate direction for the base state).
One way to mathematically think of the difference between $p_0$ and $\pi$ is that $\pi$ controls the velocity evolution 
in a way that forces the thermodynamic variables $(\rho,h,X_k)$ to evolve in a manner that is consistent with the EOS and $p_0$.\MarginPar{AJN - added this sentence; does it make sense?}

By comparing the momentum equation (\ref{eq:momentum}) to the momentum equation used in equation (2) in Paper V, we
note that we are using a formulation that enforces conservation of total energy in the
low Mach number system in the absence of external heating or viscous terms \citep{kleinpauluis,Vasil2013}.
We have previously validated this approach in modeling sub-Chandrasekhar white dwarfs using MAESTRO \citep{subChandra_II}.
We also define a one-dimensional base state density, $\rho_0 = \rho_0(r,t)$, that represents the lateral average (see Section \ref{Sec:Spatial}) of $\rho$ and is in hydrostatic equilibrium with $p_0$, i.e.,
\begin{equation}
\nabla p_0 = -\rho_0 g\eb_r, \label{eq:HSE}
\end{equation}
where $g=g(r,t)$ is the magnitude of the gravitational acceleration and $\eb_r$ is the unit vector in the outward radial direction.
Here $\beta_0$ is a density-like variable that carries background stratification, defined as
\begin{equation}
\beta_0(r,t) = \rho_0(0,t)\exp\left(\int_0^r\frac{1}{\gammaonebar p_0}\frac{\partial p_0}{\partial r'}dr'\right),
\end{equation}
where $\gammaonebar$ is the lateral average of $\Gamma_1 = d(\log p)/d(\log\rho) |_s$ (evaluated with entropy, $s$, constant).
Thermal diffusion is not discussed in this paper, but we have previously described the modifications to the original algorithm required
for implicit thermal diffusion in \cite{XRB_I}; inclusion of these effects in the new algorithm presented here is straightforward.

### 2019-08-01 MAESTROeX_I/intro.tex
fixed a couple of typos
minor changes and comments through the alg description

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR) 
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for 
flows where the Mach number is small ($\sim 0.1$ or smaller).  
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited 
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized 
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE 
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the 
pressure-projection steps.  This, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state 
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buoyancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses a predictor-corrector strategy for evolving the base state,
Strang splitting to integrate the thermodynamic variables, and a 
second-order projection method to integrate the velocity equations subject to a divergence constraint.
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising second-order accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components, 
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.  
In particular, schemes based on deferred corrections have been used to generate high-order temporal integrators for problems of reactive flow \citep{dutt2000spectral} and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state for spherical problems.
Finally, we examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries \citep{starkiller} ({\tt https://github.com/starkiller-astro}) ,
as well as AMReX ({\tt https://github.com/AMReX-Codes/amrex}).

>>>>

Previously, we developed the low Mach number astrophysical solver, MAESTRO.
MAESTRO is a block-structured, Cartesian grid finite-volume, adaptive mesh refinement (AMR)
code that has been successfully used for many years for a number of applications, detailed below.
Unlike several of the references above, MAESTRO is not an all-Mach solver, but is suitable for
flows where the Mach number is small ($\sim 0.1$ or smaller).
Furthermore, the low Mach number model in MAESTRO is specifically designed for, but not limited
to, astrophysical settings with significant atmospheric stratification.
This includes full spherical stars, as well as planar simulations of dynamics within localized
regions of a star.
The numerical methodology relies on an explicit Godunov approach for advection, a stiff ODE
solver for reactions (VODE) \citep{vode}, and multigrid-based linear solvers for the
pressure-projection steps.  Thus, the time step is limited by an advective CFL constraint based on the 
fluid velocity, not the sound speed.
Central to the algorithm are time-varying, one-dimensional stratified background (or base) state
density and pressure fields that are held in hydrostatic equilibrium.
The base state density couples to the full state solution through buoyancy terms in the momentum equation,
and the base state pressure couples to the full state solution by constraining the evolution of the
thermodynamic variables to match this pressure.
The time-advancement strategy uses Strang splitting to integrate the thermodynamic variables, a
second-order projection method to integrate the velocity subject to a divergence constraint,
and a velocity splitting scheme that uses a radially-averaged velocity to hydrodynamically evolve the base state.\MarginPar{AJN changed this sentence.}
The original MAESTRO code was developed in the pure-Fortran 90 FBoxLib software framework, whereas
MAESTROeX is developed in the C++/F90 AMReX framework \citep{AMReX,AMReX_JOSS}.

@@

In this paper, we present new algorithmic methodology that improves upon Paper V in a number of ways.
First, the overall temporal algorithm has been greatly simplified without compromising second-order accuracy.
The key design decisions were to eliminate the splitting of the velocity into average and perturbational components,
and also to replace the hydrodynamic evolution of the base state with a predictor-corrector approach.
Not only does this greatly simplify the dynamics of the base
state, but this treatment is more amenable to higher-order multiphysics coupling strategies
based on method-of-lines integration.
In particular, schemes based on deferred corrections \citep{dutt2000spectral} have been used to generate 
high-order temporal integrators for problems of reactive flow and low Mach number combustion \citep{pazner2016high,nonaka2018conservative}.
Second, we explore the effects of alternative spatial mapping routines for coupling the base state and the Cartesian grid state for spherical problems.
Finally, we examine the performance of our new MAESTROeX implementation in the new C++/F90 AMReX public software library \citep{AMReX,AMReX_JOSS}.
MAESTROeX uses MPI+OpenMP parallelism and scales well to over 10,000 MPI processes, with each MPI process supporting tens of threads.
The resulting code is publicly available on GitHub ({\tt https://github.com/AMReX-Astro/MAESTROeX}),
uses the Starkiller-Astro microphysics libraries \citep{starkiller} ({\tt https://github.com/starkiller-astro}) ,
as well as AMReX ({\tt https://github.com/AMReX-Codes/amrex}).
